---
title: 'IMT 573: Problem Set 1 - Exploring Data'
author: "Insert Your Name Here"
date: 'Due: Tuesday, October 9, 2018 at or before 11:59AM'
output: pdf_document
---

<!-- This syntax can be used to add comments that are ignored during knitting process. -->

##### Collaborators: 

##### Instructions: #####

Before beginning this assignment, please ensure you have access to R and RStudio. 

1. Download the `problemset1.Rmd` file from Canvas. Open `problemset1.Rmd` in RStudio and supply your solutions to the assignment by editing `problemset1.Rmd`. 

2. Replace the "Insert Your Name Here" text in the `author:` field with your own full name. Any collaborators must be listed on the top of your assignment. 

3. Be sure to include well-documented (e.g. commented) code chucks, figures and clearly written text chunk explanations as necessary. Any figures should be clearly labeled and appropriately referenced within the text. If you are using more than just a standard function that you found from another source, please credit the source in the comments. For example: 

```{r citing}
# code adapted from "Example: Multiplication Table"  https://www.datamentor.io/r-programming/examples/multiplication-table/


# assign num
num = 8
# use for loop to iterate 10 times
for(i in 1:10) {
print(paste(num,'x', i, '=', num*i))
}
```

4. Collaboration on problem sets is acceptable, and even encouraged, but each student must turn in an individual write-up in his or her own words and his or her own work. The names of all collaborators must be listed on each assignment. Do not copy-and-paste from other students' responses or code.

5. When you have completed the assignment and have **checked** that your code both runs in the Console and knits correctly when you click `Knit PDF`, rename the R Markdown file to `YourLastName_YourFirstName_ps1.Rmd`, knit a PDF and submit both the PDF file on Canvas.

##### Setup: #####

In this problem set you will need, at minimum, the following R packages. If you haven't yet installed them you will need to begin by using `install.packages()`

```{r Setup, message=FALSE}
# Load standard libraries
library(tidyverse)
library(nycflights13)
```

#### Problem 1: Exploring the NYC Flights Data ####

In this problem set we will use the data on all flights that departed NYC (i.e. JFK, LGA or EWR) in 2013. You can find this data in the `nycflights13` R package. 

##### (a) Importing and Inspecting Data: ##### 
Load the data and describe in a short paragraph how the data was collected and what each variable represents. Perform a basic inspection of the data and discuss what you find.

##### (b) Formulating Questions: ##### 
Consider the NYC flights data. Formulate two motivating questions you want to explore using this data. Describe why these questions are interesting and how you might go about answering them. 

##### (c) Exploring Data: ##### 

For each of the questions you proposed in Problem 1b, perform an exploratory data analysis designed to address the question. At a minimum, you should produce two visualizations related to each question. Be sure to describe what the visuals show and how they speak to your question of interest. 

##### (d) Challenge Your Results: ##### 

After completing the exploratory analysis from Problem 1c, do you have any concerns about your findings? Comment on any ethical and/or privacy concerns you have with your analysis. 
